### 2019-06-16 Manuscripts/cog-psych/post-revision/post-cogpsych.tex
pass through

% ------- ABSTRACT PAGE ------- %
\doublespacing
\section*{Abstract}
The ability to generate new concepts and ideas is among the most fascinating
aspects of human cognition, but we do not have a strong understanding of the
cognitive processes and representations underlying concept generation. In this
paper, we study the generation of new categories using the computational and
behavioral toolkit of traditional artificial category learning. Previous work in
this domain has focused on how the statistical structure of known categories
generalizes to generated categories, overlooking the extent that the contrast
between the known and generated categories is a factor. We report three
experiments demonstrating that contrast between what is known and what is
created is of fundamental importance for categorization. We propose two novel
approaches to modeling category contrast: one focused on exemplar dissimilarity
and another on the representativeness heuristic. Our experiments and
computational analyses demonstrate that both models capture different aspects of
contrast's role in categorization.

@@

Creating and learning new ideas is one of the most fascinating and important
human capabilities. For example, cell phones and smart phones are newly created
categories of objects that young and old adults learned and have become reliant
on in the last few decades. Computational researchers have identified
generating novel objects and objectives as one of the defining, and most
difficult to formalize, characteristics of intelligent life
\citep{lake2017,lehman2011,taylor2016}. Yet, the cognitive mechanisms that
enable people to innovate and learn new innovations are not well understood and
are understudied by scientists. In part, this is due to the inherent difficulty
of designing and conducting experiments that test these capabilities. How would
a scientist devise an experiment that induces a participant to create and learn
new categories that are as interesting as cell phones and smart phones? As a
step towards these goals, in this article we investigate the expectations that
people have when they learn and generate new categories using computational
models and behavioral experiments.

Generating and learning new concepts are, however, not altogether different from
the types of behaviors typically studied in cognitive psychology laboratories.
In particular, generating a member of a novel class can be considered a `special
case' application of existing category knowledge
\citep{kemp2014taxonomy,kurtz2015human}. Research in categorization typically
focuses on what properties of a category affect human learning of an object's
category given its features \citep{shepard1961learning,kurtz2013human}, or the
prediction of an object's unobserved features given some of its other features
and/or its category \citep{markman2003category}. The generation of members of a
new category consists of inferring {\em all} features for a {\em novel} category
label. Thus, we can make progress formalizing the processes involved in category
generation by extending theories of categorization to this case. Then, to ensure
that these principles reflect basic principles of categorization and are not
just specific to category generation, we can examine what factors of the
generated categories affect the ease at which people learn the category.

Previous work in category learning and generation has established that people
are highly sensitive to the structural properties of categories, such as
correlations between the features of category members and the relation between
items within the same category and those in different categories
\citep{roschmervis1975,regier2007,shepard1961learning,smith1995creative}.
Inspired by this work, previous research on the topic of category generation has
explored a similar principle: People tend to create new categories that have
similar {\em statistical regularities} as previously learned categories
\citep{jern2013probabilistic,ward1994structured}. Although this is an important
characteristic of generating new categories, it cannot be the only one. Taken to
the extreme, the best ``new'' category in terms of having the same statistical
regularities to other categories would be identical to a known category that is
representative of the domain (and thus, not new at all). Contrast from other
known categories should play a role. Further, scientists in other fields, such
as marketing \citep{berger16} and sociology \citep{rogers03}, highlight the
critical importance of contrast in the creation of new ideas.

@@

Although implicitly assumed in some work, this constraint has been
overlooked in previous research: To our knowledge, there has not been any
systematic investigation addressing how generated categories {\em differ} from
what is already known. Although the idea of category contrast is ubiquitous
throughout the categorization literature, and extends to a variety of other
fields \citep[e.g., color;][]{regier2007}, the idea that a new category should
be ``different'' is vague, as there are many ways it could be different from a
previously observed category.

@@

The outline of the article is as follows. First we describe previous
computational formalizations of theories of category generation and empirical
work investigating them. We then present two hypotheses for how contrast might
affect categorization and formalize them in computational models. The first is a
novel exemplar model, which is designed to generate categories that
systematically differ from existing categories in the domain. The second assumes
that the goal of category generation is to create representative samples of the
opposite of the observed categories. We present two experiments demonstrating
strong and systematic effects of category contrast on concept generation, and we
qualitatively and quantitatively analyze the performance of each model in
capturing human category generation. A third experiment is performed to
highlight the differences between the two contrast models. We conclude with a
discussion of the implications of our results for categorization and directions
for future work.

@@

\cite{jern2013probabilistic} additionally tested a ``copy-and-tweak'' model that
broadly resembles the earlier ``Path of Least Resistance'' account. The core
proposal is that participants generate new items by copying stored examples from
memory and tweaking them to generate something new. The copy-and-tweak model
differs from the Path of Least Resistance account in that it notably omits the
hierarchical organization of categories, as well as selectivity to which
features are changed \citep[both of which are factors in the Path of Least
Resistance account;][]{ward2002role}. Instead, their copy-and-tweak model
corresponds to a direct exemplar-similarity approach
\citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}, generating new items
according to their similarity to known members of the target category. The
copy-and-tweak model provided a poor account of their results, as the
experiments devised by \citeauthor{jern2013probabilistic} were specifically
designed to challenge it. Its application is notable as a first step
toward understanding theories developed in the creative cognition literature
using well-known formal approaches from the categorization literature.

@@

Prior work on category generation has explored only one factor: distributional
correspondences between learned and generated categories. As a result, most of
the computational, theoretical, and empirical efforts have been focused on
explaining those effects. In this paper, we investigate another important
factor: category contrast. To generate a novel concept, individuals must produce
something that is in some capacity {\em different} from what they already know.
As a consequence, we propose that contrast should be a primary constraint that
underlies people's expectations about how a set of categories are organized:
Categories should be different from each other.

Although it is evident that people are {\em capable} of creating new concepts
and categories, it is not entirely clear how new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but using a broader class of standard hierarchical Bayesian models without any
additional features \citep[e.g., ][]{griffithsscn07,kemp07}. Many of these
models assume that at some point of the latent generative process the same
underlying distribution generates all of the categories and thus, any
differences between categories are due to {\em noise} and should not be {\em
  systematic}. If the goal of creating a new category given others in a domain
is generating from the posterior predictive distribution then the best a
standard hierarchical Bayesian model without a notion of contrast built into the
model can do at capturing contrast is to assume that the new category is placed
uniformly at random over stimulus space. But, this defeats the purpose of a
hierarchy as it is ignored when determining a new category location!\footnote{It
  is plausible that some hierarchical Bayesian models could be created that
  generates categories different from each other. However, this model would not
  be a standard application or extension of most pre-existing hierarchical
  Bayesian models. The generative process would need to include a component that
  presumes contrast.} Note that this does not mean location information and
contrast cannot be captured using hierarchical Bayesian models. In fact, a model
instantiating contrast as representativeness is a hierarchical Bayesian model.
But, it requires an additional factor, which in our case is having a different
goal than predicting the next category.

@@

Although, as an exemplar model, one of PACKER's proposals is people represent
categories in terms of a collection of stored exemplars, we did not pick it
assuming it is the correct model of human categorization. The choice to develop
PACKER within an exemplar framework reflects the facts that exemplar models have
been throughly evaluated, are strongly theoretically motivated, and dominate
much of the theoretical and empirical work in categorization. The focus of our
work with PACKER concerns the dual constraints of within- and between-class
similarity; it is not difficult to imagine how such constraints may be
instantiated using alternative frameworks (e.g.,
\citealp{kurtz2007divergent,love2004sustain,smith2000thirty}; for a review of
categorization models, see \citealp{pothoswills2011}).

@@

\subsubsection{Relation Between PACKER and Copy-And-Tweak}
\label{section:copytweak-packer} The PACKER model is similar to the
copy-and-tweak model reported by \cite{jern2013probabilistic}: Both models are
exemplar-based, and both models generate new items according to their similarity
to known members of the target class. However, PACKER diverges from the
copy-and-tweak model by including a contrast mechanism, enabling generation
according to dissimilarity to members of opposing categories. As a consequence,
copy-and-tweak can be realized as a parameterization of the PACKER model that is
insensitive to category contrast. Specifically, when $\theta_c = 0$ and
$\theta_t > 0$ (see Figure \ref{fig:packer-examples}, panel B), $f(x_j) = 0$ for
$x_j$ belonging to contrast categories; thus, PACKER is not influenced by these
items, and is mathematically equivalent to a copy-and-tweak approach.

In this paper, we report simulations using this copy-and-tweak model. This model
fits within the exemplar-based category generation framework defined above, and
is a continuous-dimension adaptation of the model tested by
\cite{jern2013probabilistic}. By formalizing a model family where PACKER and
copy-and-tweak are different parameterizations of models within the same
framework, the comparison between PACKER and copy-and-tweak provides a test of
the explanatory value of the contrast mechanism based on exemplar dissimilarity:
The account provided by copy-and-tweak will only equal that of PACKER if the
contrast mechanism does not offer an advantage (i.e., if $\theta_c > 0$
significantly improves model fits). Note that the purpose of the article is to
explore and formally analyze the role of contrast in categorization and thus, we
leave extending PACKER to incorporate distributional factors for future work.

@@

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they resemble the target
category and are thus more distinct from and less similar to other categories. We
adopt the formalization given by \cite{tenenbaum01}, where the
representativeness $R(y,h)$ of an item $y$ (or in our case, an exemplar) is the
relative amount of evidence that is provided by $y$ for a given hypothesis $h$
in a space of hypotheses $\mathcal{H}$ in contrast to all other hypotheses
$\mathcal{H}^c=\mathcal{H} \setminus \{h\}$:
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{1-p(x|\mathcal{H}^c)} = \log \dfrac{p(x|h)}{\sum_{h'\in H'}{p(x|h')p(h')}} . 
\label{eq:representativeness}
\end{equation}
In Equation \ref{eq:representativeness}, $p(h')$ is the prior probability
distribution on the hypothesis space that excludes $h$. Note that when there are
only two hypotheses (e.g. two categories) involved in the domain, this prior
takes the value of $1$ since there is ever only one alternative hypothesis.

@@

At this point, both the representativeness model and the hierarchical Bayesian
model from \cite{jern2013probabilistic} are largely identical. However, both
models diverge in their computation of exemplar generation probabilities. While
the original hierarchical Bayesian model produces novel exemplars with
probabilities proportional to the multivariate normal likelihood $p(y|h)$, the
representativeness model generates new exemplars according to their
representativeness $R(y,h)$. In practice, the probability of generating a
particular candidate $y$ is obtained using an Exponentiated
\citet{luce1977choice} choice rule:
\begin{equation}
  p(y) = \dfrac
    {\exp( \theta \cdot R(y,h))}
    {\sum_k \exp(\theta  \cdot R(y_k,h_k))} .
\label{eq:representative-choice}
\end{equation}
where $\theta$ is a freely estimated response determinism parameter (constrained
such that $\theta \geq 0$).

@@

\subsection{Synopsis and Prognosis}

Research on categorization and category generation has focused on the finding
that generated categories tend to possess distributional commonalities with
known categories. However, a fundamental task of categorization is to create
something {\em new} (i.e., different from what is already known). The manner in
which generated categories differ from known ones is, nonetheless, poorly
understood: Existing theories do not make strong predictions about how generated
concepts should systematically differ from existing ones. Above, we introduced
two novel approaches to formalizing the roles of similarity and contrast in
category generation.

In the sections below, we present two experiments demonstrating systematic
effects of category contrast on category generation inspired by factors
influencing how the contrast models generate new categories. Our experiments are
based on \cite{jern2013probabilistic}'s paradigm, which is a straightforward
translation of the traditional artificial classification paradigm to the task of
category generation: Participants are first exposed to a single,
experimenter-defined category, and are then asked to generate members of a new
category. We then report quantitative analyses of our results comparing the exemplar dissimilarity
and the representativeness accounts of our results to that of the hierarchical
Bayesian and copy-and-tweak models developed by \cite{jern2013probabilistic}.
After analyzing the category generation studies, we present a third experiment to
distinguish between our two conceptualizations of contrast.

@@

Interestingly, the XOR condition also offers a dissociation between the roles of
category contrast and the emulation of distributional structure:
widely-distributed, positively-correlated categories would need to lie along the
positive diagonal of the space (that is the only place they ``fit''), which is
already occupied by the experimenter-defined category. Thus, if contrast plays a
role, exemplars in the generated categories of participants in the XOR condition
may not be positively correlated -- they may be negatively correlated instead.
In this case, contrast and statistical regularities would interact, which would
be inconsistent with prevailing theories of category generation
\citep{jern2013probabilistic}.

@@

A secondary goal of this experiment was to examine whether we replicate the
classic result that generated categories often possess the same distributional
properties as previously-known categories. Given the increased emphasis of replication within psychology \citep{zwaan2018making}, it is important as it serves as a conceptual replication of \citet{jern2013probabilistic}. For each generated category, we
computed the category range along each axis (X, Y), as well as the correlation
between features. These data, shown in Figure \ref{fig:e1-statsboxes}, reveal
broad individual differences: Within each condition, participants generated
categories spanning the entire X- and Y- axis as well as categories that spanned
very little along each. Likewise, in each condition participants generated
categories possessing strongly positive, neutral, and strongly negative
correlations between the dimensions. Comparing the distributional statistics
between conditions yields a broad yet, as we will see, misleading replication of
the classic effect.

@@

\subsection{Discussion} In Experiment 1 we sought to extend our analysis of the
\cite{jern2013probabilistic} data by evaluating the influence of category
contrast on category generation, given qualitatively different types of categories. We found strong evidence for effects of category contrast in each
condition: Participants were more likely to generate stimuli that are more
distant from (i.e., less similar to) members of a previously-learned category,
and members of participant-generated categories tended to be more similar to one
another than to members of previously-learned categories. We also partially
replicated the classic finding that the distributional structure of generated
categories reflects that of previously learned categories
\citep{jern2013probabilistic,ward1994structured}: Members of generated
categories were more widely distributed along dimensions which were widely
distributed in the experimenter-defined category.

@@

However, our ability to detect differences in Y-Axis range using a standard
$t$-test between the conditions is, in this case, diminished due to the
non-normality of the data (Shapiro-Wilk normality test $W=0.77, p < .001$ for
the Middle condition and $W=0.85, p < .001$ for the Bottom condition). Figure
\ref{fig:e2-yranges} depicts the Y-axis position of the exemplars generated
within each participant's category. Further, the median produced Y-Axis range is much smaller in the Bottom than Middle condition, whereas they are essentially identical in X-Axis range. The categories are sorted by overall range,
then by condition assignment. These data reveal that there were nearly as many
participants who generated categories spanning the entire Y-axis as those who
generated categories spanning almost none of the Y-axis. The non-normality of
the Y-axis range distributions thus requires that we use a different approach to
addressing the experiment's main question.

@@

Table \ref{table:global-model-fits} contains the model fits. Due to the uneven number of fitted parameters among the models, we
compare the model fits using the Akaike Information Criterion
\citep[AIC;][]{akaike1974new}, where smaller values correspond to better fits
(discounted by model complexity as measured by the number of parameters). The
same qualitative results were obtained with alternative model comparison metrics
(e.g., BIC, \citealp{schwarz1978estimating}; $AIC_C$,
\citealp{hurvich1989regression}). In addition to AIC, table \ref{table:global-model-fits} contains the corresponding log-likelihood ($L$) and the best-fitting
parameter values. These results reveal strong model differentiation: both
contrast models (PACKER and the representativeness model) achieved far better
fits compared to their non-contrast counterparts: copy-and-tweak and the
hierarchical Bayesian model respectively. Interestingly between the contrast
models, the hierarchical model (representativeness) outperformed the
exemplar-based theory (PACKER), whereas between the non-contrast models, the
reverse is observed. Specifically, here the exemplar-based model
(copy-and-tweak) performed somewhat better than the hierarchical Bayesian model.

@@

Thus, it is possible that the failure of the classical models is simply due to
their inability to explain each participant's first trial (generating the first
item in the `Beta' category). To ensure this is not driving our results, we
conducted an identical set of simulations as above, excluding this trial
(leaving 915 responses in the dataset): Again, the representativeness model ($L
= -3286$, $AIC = 6580$) and PACKER ($L = -3377$,
$AIC = 6759$) achieved better fits than the copy-and-tweak ($L = -3564$,
$AIC = 7132$) and hierarchical Bayesian ($L = -3597$, $AIC = 7201$) models.

@@

For both the hierarchical Bayesian and the representativeness models, the prior
domain covariance matrix $\Sigma_0$ can be used to explain the generation of
row-like and column-like categories. This covariance matrix specifies the amount
of variance assumed along each dimension (as well as the dimensional
correlations) across the domain of categories. The covariance matrix for a newly
generated category, $\Sigma_B$, is based on the assumed $\Sigma_0$ as well as
the distributions of previously learned categories (see Appendix
\ref{ap:hsampling-definition}). Thus, the importance of each feature can be
coded into $\Sigma_0$ to alter the dimensional variance of generated categories.
Because the hierarchical Bayesian model possesses no mechanism to account for
category contrast, the model is most likely to generate new items that are
similar to known examples of the target category with no regard for how
different it is to the contrast category. However, the representativeness model
predicts that new exemplars should provide more relative evidence to the `Beta'
category, accounting for the tendency of row-like and column-like profiles
occupying the edge of the feature space.

@@

[Maybe work on transition if there's time]

>>>>

% ------- ABSTRACT PAGE ------- %
\doublespacing
\section*{Abstract}
The ability to generate new concepts and ideas is among the most fascinating
aspects of human cognition, but we do not have a strong understanding of the
cognitive processes and representations underlying concept generation. In this
paper, we study the generation of new categories using the computational and
behavioral toolkit of traditional artificial category learning. Previous work in
this domain has focused on how the statistical structure of known categories
generalizes to generated categories, overlooking whether (and if so, how) contrast
between the known and generated categories is a factor. We report three
experiments demonstrating that contrast between what is known and what is
created is of fundamental importance for categorization. We propose two novel
approaches to modeling category contrast: one focused on exemplar dissimilarity
and another on the representativeness heuristic. Our experiments and
computational analyses demonstrate that both models capture different aspects of
contrast's role in categorization.

@@

Creating new ideas is one of the most fascinating and important
human capabilities. For example, cell phones and smart phones are newly created
categories of objects that young and old adults learned and have become reliant
on in the last few decades. Computational researchers have identified
generating novel objects and objectives as one of the defining, and most
difficult to formalize, characteristics of intelligent life
\citep{lake2017,lehman2011,taylor2016}. Yet, the cognitive mechanisms that
enable people to innovate are not well understood and
are understudied by scientists. In part, this is due to the inherent difficulty
of designing and conducting experiments that test these capabilities. How would
a scientist devise an experiment that induces a participant to create 
new categories that are as interesting as cell phones and smart phones? Devising such an experiment is still beyond our capability. Instead, we take a step towards this goal, and investigate the fundamental constraints and expectations that people have when they generate new categories.

% we do have tools to study the fundamental constraints on category . As a
%step towards these goals, in this article we investigate the expectations that
%people have when they generate new categories using computational
%models and behavioral experiments.

Creating new concepts are, however, not altogether different from
the types of behaviors typically studied in cognitive psychology laboratories.
In particular, generating a member of a novel class can be considered a `special
case' application of existing category knowledge
\citep{kemp2014taxonomy,kurtz2015human}. Research in categorization typically
focuses on what properties of a category affect human learning of an object's
category given its features \citep{shepard1961learning,kurtz2013human}, or the
prediction of an object's unobserved features given some of its other features
and/or its category \citep{markman2003category}. The generation of members of a
new category consists of inferring {\em all} features for a {\em novel} category
label. Thus, we can make progress formalizing the processes involved in category
generation by extending theories of categorization to this case. This will also serve to test the breadth of the explanatory power of categorization models to tasks beyond typical category learning.

Previous work in category learning and generation has established that people
are highly sensitive to the structural properties of categories, such as
correlations between the features of category members and the relation between
items within the same category and those in different categories
\citep{roschmervis1975,regier2007,shepard1961learning,smith1995creative}.
Inspired by this work, previous research on the topic of category generation has
explored a similar principle: People tend to create new categories that have
similar {\em statistical regularities} as previously learned categories
\citep{jern2013probabilistic,ward1994structured}. Although this is an important
characteristic of generating new categories, it cannot be the only one. Taken to
the extreme, the best ``new'' category in terms of having the same statistical
regularities to other categories would be identical to a known category that is
representative of the domain (and thus, not new at all). Contrast from other
known categories should play a role. Indeed, scientists in other fields, such
as marketing \citep{berger16} and sociology \citep{rogers03}, highlight the
critical importance of contrast in the creation of new ideas.

@@

Although implicitly assumed in some work, this constraint has been
overlooked in previous research: To our knowledge, there has not been any
systematic investigation addressing how generated categories {\em differ} from
what is already known. Although the idea of category contrast is discussed
throughout the categorization literature, and extends to a variety of other
fields \citep[e.g., color;][]{regier2007}, the idea that a new category should
be ``different'' is vague, as there are many ways it could be different from a
previously observed category.

@@

The outline of the article is as follows. First we describe previous
computational formalizations of theories of category generation and empirical
work investigating them. We then present two hypotheses for how contrast might
affect categorization and formalize them in computational models. The first is a
novel exemplar model, which is designed to generate categories that
systematically differ from existing categories in the domain. The second assumes
that the goal of category generation is to create representative samples of the
opposite of the observed categories. We present two experiments demonstrating
strong and systematic effects of category contrast on concept generation, and we
qualitatively and quantitatively analyze the behavioral results and the performance of each model in
capturing human category generation. A third experiment is performed to
highlight the differences between the two contrast models. We conclude with a
discussion of the implications of our results for categorization and directions
for future work.

@@

\cite{jern2013probabilistic} additionally tested a ``copy-and-tweak'' model that
broadly resembles the earlier ``Path of Least Resistance'' account. The core
proposal is that participants generate new items by copying stored examples from
memory and tweaking them to generate something new. The copy-and-tweak model
differs from the Path of Least Resistance account in that it notably omits the
hierarchical organization of categories, as well as selectivity to which
features are changed \citep[both of which are factors in the Path of Least
Resistance account;][]{ward2002role}. Instead, their copy-and-tweak model
corresponds to a direct exemplar-similarity approach
\citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}, generating new items
according to their similarity to known members of the target category. The
copy-and-tweak model provided a poor account of their results, as the
experiments devised by \citeauthor{jern2013probabilistic} were specifically
designed to challenge it.

@@

%Prior work on category generation has explored only one factor: distributional
%correspondences between learned and generated categories. As a result, most of
%the computational, theoretical, and empirical efforts have been focused on
%explaining those effects. In this paper, we investigate another important
%factor: category contrast. To generate a novel concept, individuals must produce
%something that is in some capacity {\em different} from what they already know.
%As a consequence, we propose that contrast should be a primary constraint that
%underlies people's expectations about how a set of categories are organized:
%Categories should be different from each other.

Although people are {\em capable} of creating 
new categories, it is not entirely clear whether (and if so, how) new concepts are systematically
made different from what is already known. The hierarchical Bayesian model
developed by \cite{jern2013probabilistic} assumes that differences between
observed and generated categories are only due to random variation. The model
assumes that generated categories are sampled from the same underlying domain
distribution as observed categories, and will thus share a common distributional
structure. The model does not make predictions about the {\em location} of the
category within the domain (the perceptual instantiation of category members).
Under a strict interpretation of their model, given knowledge of a single
category within the domain, the most probable new category to be generated is
located in {\em exactly} the same location and possesses an identical
distributional structure. This is not an issue with their model specifically,
but using a broader class of standard hierarchical Bayesian models without any
additional features \citep[e.g., ][]{griffithsscn07,kemp07}. Many of these
models assume that at some point of the latent generative process the same
underlying distribution generates all of the categories and thus, any
differences between categories are due to {\em noise} and should not be {\em
  systematic}. If the goal of creating a new category given others in a domain
is generating from the posterior predictive distribution then the best a
standard hierarchical Bayesian model without a notion of contrast built into the
model can do at capturing contrast is to assume that the new category is placed
uniformly at random over stimulus space. But, this defeats the purpose of a
hierarchy as it is ignored when determining a new category location!\footnote{It
  is plausible that some hierarchical Bayesian models could be created that
  generates categories different from each other. However, this model would not
  be a standard application or extension of most pre-existing hierarchical
  Bayesian models. The generative process would need to include a component that
  presumes contrast, which is precisely the factor of study in this article.} Note that this does not mean location information and
contrast cannot be captured using hierarchical Bayesian models. In fact, a model
instantiating contrast as representativeness is a hierarchical Bayesian model.
But, it requires an additional factor, which in our case is having a different
goal than predicting the next category.

@@

Although, as an exemplar model, one of PACKER's proposals is people represent
categories in terms of a collection of stored exemplars, we did not pick it
assuming it is the correct model of human categorization. The choice to develop
PACKER within an exemplar framework reflects the facts that exemplar models have
been throughly evaluated, are strongly theoretically motivated, and dominate
much of the theoretical and empirical work in categorization. The focus of our
work with PACKER concerns how contrast may be captured through the dual constraints of within- and between-class
similarity; it is not difficult to imagine how such constraints may be
instantiated using alternative frameworks (e.g.,
\citealp{kurtz2007divergent,love2004sustain,smith2000thirty}; for a review of
categorization models, see \citealp{pothoswills2011}).

@@

\subsubsection{Relation Between PACKER and Copy-And-Tweak}
%: Both models are
%exemplar-based, and both models generate new items according to their similarity
%to known members of the target class. However, PACKER diverges from the
%copy-and-tweak model by including a contrast mechanism, enabling generation
%according to dissimilarity to members of opposing categories. As a consequence,
%copy-and-tweak can be realized as a parameterization of the PACKER model that is
%insensitive to category contrast. Specifically, when $\theta_c = 0$ and
%$\theta_t > 0$ (see Figure \ref{fig:packer-examples}, panel B), $f(x_j) = 0$ for
%$x_j$ belonging to contrast categories; thus, PACKER is not influenced by these
%items, and is mathematically equivalent to a copy-and-tweak approach.

\label{section:copytweak-packer} The PACKER model is similar to the copy-and-tweak model reported by \cite{jern2013probabilistic}.
An advantage of formulating copy-and-tweak as a special case parameterization of PACKER. In this paper, we report simulations using this copy-and-tweak model. This model
fits within the exemplar-based category generation framework defined above, and
is a continuous-dimension adaptation of the model tested by
\cite{jern2013probabilistic}. By formalizing a model family where PACKER and
copy-and-tweak are different parameterizations of models within the same
framework, the comparison between PACKER and copy-and-tweak provides a test of
the explanatory value of the contrast mechanism based on exemplar dissimilarity:
The account provided by copy-and-tweak will only equal that of PACKER if the
contrast mechanism does not offer an advantage (i.e., if $\theta_c > 0$
significantly improves model fits). Note that the purpose of the article is to
explore and formally analyze the role of contrast in categorization and thus, we
leave extending PACKER to incorporate distributional factors for future work.

@@

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they resemble the target
category and are thus more distinct from and less similar to other categories. We
adopt the formalization given by \cite{tenenbaum01}, where the
representativeness $R(y,h)$ of an item $y$ (or in our case, an exemplar) is the
relative amount of evidence that is provided by $y$ for a given hypothesis $h$
in a space of hypotheses $\mathcal{H}$ in contrast to all other hypotheses
$\mathcal{H}^c=\mathcal{H} \setminus \{h\}$:
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{1-p(x|\mathcal{H}^c)} = \log \dfrac{p(x|h)}{\sum_{h'\in \mathcal{H}^c}{p(x|h')p(h')}} . 
\label{eq:representativeness}
\end{equation}
In Equation \ref{eq:representativeness}, $p(h')$ is the prior probability
distribution on the hypothesis space that excludes $h$. Note that when there are
only two hypotheses (e.g. two categories) involved in the domain, this prior
takes the value of $1$ since there is ever only one alternative hypothesis.

@@

At this point, both the representativeness model and the hierarchical Bayesian
model from \cite{jern2013probabilistic} are largely identical. However, both
models diverge in their computation of exemplar generation probabilities. While
the original hierarchical Bayesian model produces novel exemplars with
probabilities proportional to the multivariate normal likelihood $p(y|h)$, the
representativeness model generates new exemplars according to their representativeness $R(y,h)$ as formulated in Equation \ref{eq:representativeness}. In practice, the probability of generating a
particular candidate $y$ is obtained using an Exponentiated
choice rule \citep{luce1977choice}:
\begin{equation}
  p(y) = \dfrac
    {\exp( \theta \cdot R(y,h))}
    {\sum_k \exp(\theta  \cdot R(y_k,h_k))} .
\label{eq:representative-choice}
\end{equation}
where $\theta$ is a freely estimated response determinism parameter (constrained
such that $\theta \geq 0$).

@@

%\subsection{Synopsis and Prognosis}
%
%Research on categorization and category generation has focused on the finding
%that generated categories tend to possess distributional commonalities with
%known categories. However, a fundamental task of categorization is to create
%something {\em new} (i.e., different from what is already known). The manner in
%which generated categories differ from known ones is, nonetheless, poorly
%understood: Existing theories do not make strong predictions about how generated
%concepts should systematically differ from existing ones. Above, we introduced
%two novel approaches to formalizing the roles of similarity and contrast in
%category generation.
%
%In the sections below, we present two experiments demonstrating systematic
%effects of category contrast on category generation inspired by factors
%influencing how the contrast models generate new categories. Our experiments are
%based on \cite{jern2013probabilistic}'s paradigm, which is a straightforward
%translation of the traditional artificial classification paradigm to the task of
%category generation: Participants are first exposed to a single,
%experimenter-defined category, and are then asked to generate members of a new
%category. We then report quantitative analyses of our results comparing the exemplar dissimilarity
%and the representativeness accounts of our results to that of the hierarchical
%Bayesian and copy-and-tweak models developed by \cite{jern2013probabilistic}.
%After analyzing the category generation studies, we present a third experiment to
%distinguish between our two conceptualizations of contrast.

@@

Interestingly, the XOR condition also offers a dissociation between the roles of
category contrast and the emulation of distributional structure:
widely-distributed, positively-correlated categories would need to lie along the
positive diagonal of the space (that is the only place they ``fit''), which is
already occupied by the experimenter-defined category. Thus, if contrast plays a
role, exemplars in the generated categories of participants in the XOR condition
may not be positively correlated -- they may not be correlated, or perhaps even be negatively correlated.
In this case, contrast and statistical regularities would interact, which would
be inconsistent with prevailing theories of category generation
\citep{jern2013probabilistic}.

@@

A secondary goal of this experiment was to examine whether we replicate the
classic result that generated categories often possess the same distributional
properties as previously-known categories. Given the increased importance of replication within psychology \citep{zwaan2018making}, it is important as it serves as a conceptual replication of \citet{jern2013probabilistic}. For each generated category, we
computed the category range along each axis (X, Y), as well as the correlation
between features. These data, shown in Figure \ref{fig:e1-statsboxes}, reveal
broad individual differences: Within each condition, participants generated
categories spanning the entire X- and Y- axis as well as categories that spanned
very little along each. Likewise, in each condition participants generated
categories possessing strongly positive, neutral, and strongly negative
correlations between the dimensions. Comparing the distributional statistics
between conditions yields a broad yet, as we will see, misleading replication of
the classic effect.

@@

\subsection{Discussion} In Experiment 1 we evaluated the influence of category
contrast on category generation, given qualitatively different types of categories. We found strong evidence for effects of category contrast in each
condition: Participants were more likely to generate stimuli that are more
distant from (i.e., less similar to) members of a previously-learned category,
and members of participant-generated categories tended to be more similar to one
another than to members of previously-learned categories. We also partially
replicated the classic finding that the distributional structure of generated
categories reflects that of previously learned categories
\citep{jern2013probabilistic,ward1994structured}: Members of generated
categories were more widely distributed along dimensions which were widely
distributed in the experimenter-defined category.

@@

However, our ability to detect differences in Y-Axis range using a standard
$t$-test between the conditions is, in this case, diminished due to the
non-normality of the data (Shapiro-Wilk normality test $W=0.77, p < .001$ for
the Middle condition and $W=0.85, p < .001$ for the Bottom condition). Figure \ref{fig:e2-yranges} depicts the Y-Axis range and Y-Axis position of exemplars generated by each participant. The categories are sorted by overall range, and then colored by training condition. These data reveal that there were nearly as many
participants who generated categories spanning the entire Y-axis as those who
generated categories spanning almost none of the Y-axis. Indeed, the median produced Y-Axis range is much smaller in the Bottom than Middle condition, whereas they are essentially identical in X-Axis range. The non-normality of the Y-axis range distributions thus requires that we use a different analytic approach to
addressing the experiment's main question.

@@

Table \ref{table:global-model-fits} contains the model fits. Due to the uneven number of fitted parameters among the models, we
compare the model fits using the Akaike Information Criterion
\citep[AIC;][]{akaike1974new}, where smaller values correspond to better fits
(discounted by model complexity as measured by the number of parameters). The
same qualitative results were obtained with alternative model comparison metrics
(e.g., BIC, \citealp{schwarz1978estimating}; $AIC_C$,
\citealp{hurvich1989regression}). In addition to AIC, Table \ref{table:global-model-fits} contains the corresponding log-likelihood ($L$) and the best-fitting
parameter values. These results reveal strong model differentiation: both
contrast models (PACKER and the representativeness model) achieved far better
fits compared to their non-contrast counterparts: copy-and-tweak and the
hierarchical Bayesian model respectively. Interestingly between the contrast
models, the hierarchical model (representativeness) outperformed the
exemplar-based theory (PACKER), whereas between the non-contrast models, the
reverse is observed. Specifically, here the exemplar-based model
(copy-and-tweak) performed somewhat better than the hierarchical Bayesian model.

@@

Thus, it is possible that the failure of the classical models is simply due to
their inability to explain each participant's first trial (generating the first
item in the `Beta' category). To ensure this is not driving our results, we
conducted an identical set of simulations as above, excluding the first trial
(leaving 915 responses in the dataset): Again, the representativeness model ($L
= -3286$, $AIC = 6580$) and PACKER ($L = -3377$,
$AIC = 6759$) achieved better fits than the copy-and-tweak ($L = -3564$,
$AIC = 7132$) and hierarchical Bayesian ($L = -3597$, $AIC = 7201$) models.

@@

For both the hierarchical Bayesian and the representativeness models, the prior
domain covariance matrix $\Sigma_0$ can be used to explain the generation of
row-like and column-like categories. This covariance matrix specifies the amount
of variance assumed along each dimension (as well as the correlations between dimensions) across the domain of categories. The covariance matrix for a newly
generated category, $\Sigma_B$, is based on the assumed $\Sigma_0$ as well as
the distributions of previously learned categories (see Appendix
\ref{ap:hsampling-definition}). Thus, the importance of each feature can be
coded into $\Sigma_0$ to alter the dimensional variance of generated categories.
Because the hierarchical Bayesian model possesses no mechanism to account for
category contrast, the model is most likely to generate new items that are
similar to known examples of the target category with no regard for how
different it is to the contrast category. However, the representativeness model
predicts that new exemplars should provide more relative evidence to the `Beta'
category, accounting for the tendency of row-like and column-like profiles
occupying the edge of the feature space.

@@

%[Maybe work on transition if there's time]
